Improved Hierarchical Clustering Through Nesting

### 2015-06-02 icdm/submission.tex
Edits

\begin{abstract}
Hierarchical clustering organizes large data into smaller, more informative groups and offers a global view of the data structure by producing a hierarchy of clusters starting from individual data points and merging towards the cluster of all data. Typically, hierarchical clustering is done by greedily either merging or dividing clusters, called agglomerative and divisive clustering, respectively. Lin et al.~\cite{Lin} proposed a nesting framework that produces a hierarchal clustering from partitional clusterings for every possible number of clusters. This algorithm was introduced as an approximation algorithm for the hierarchical $k$-median problem and guarantees the resulting hierarchical layers are a limited distance from the original partitional clusters. However, this algorithm has not been previously placed in a clustering context. We test this algorithm using various methods for producing the partitional clusterings against the traditional hierarchical clustering methods and show both the approximation ratio and FScore measure improve significantly. We also propose a heuristic for this nesting framework which improves the run time and only suffers a slight decrease in approximation and FScore.
\end{abstract}

\section{Introduction and Notation}
As the data we have access to grows in variety and size and becomes cheaper to store, we need methods to efficiently learn from this data. 
Partitioning data (clustering) into similar groups (clusters) is one of the key tools to learn about a set of unlabeled data such as a set of customers who have visited a website or a set of handwritten documents.
Clustering in these situations helps us learn underlying structure and classifications of our data, and it allows large data sets to be presented in a simplified and compressed view \cite{Jain}.

@@

Clustering algorithms date back more than 50 years, and there have been many variations introduced since then \cite{Jain}. However, clustering algorithms are often formulated as NP-hard problems with non-convex objective functions, which are inherently hard to solve \cite{Jain}.
One common objective function for partitional clustering is to minimize the sum of distances from each point to the median point of its cluster, called $k$-median or $k$-medoids clustering. More specifically, suppose we are given a set $X$ and a distance (dissimilarity) metric $d(i,j)$ for all $i,j \in X$. The goal is to find a set of centers $S \subseteq X$ such that $|S| \leq k$ so as to minimize 
\[ f(S) = \sum_{ i\in X} \min_{j \in S} d(i,j). \]
The clustering is obtained from the set of centers $S$ by assigning each data point to its closest center.  An $\alpha$-approximation algorithm for this problem produces a set $S$ in polynomial time such that $f(S) \leq \alpha \cdot f(O)$ for all $O \subseteq X$ with $|O| = k$.

\begin{figure*}[!]
\centerline{\subfloat[Partitional Clustering]{\includegraphics[width =1.75in]{partitional_clustering}
\label{fig_part}}
\hfil
\subfloat[Hierarchical Clustering]{\includegraphics[width=2.5in]{hier_clustering}
\label{fig_hier}}}
\caption{A partitional clustering for $k=3$, where the colors represent the different partitions, and a dendrogram representation of a hierarchical clustering of the data where edges represent merges of clusters and the clustering on the third level down (represented by the red line) corresponds to the partitional clustering on the left.}
\end{figure*}

@@

\subsection{Partitional Clustering}
Arguably the most common partitional clustering method is $k$-means, which chooses $k$ cluster centers points in space to minimize the sum of squared distances from each point to its closest cluster center \cite{Jain}.  $k$-means is usually solved using Lloyd's algorithm (also referred to simply as the $k$-means algorithm), which iteratively greedily updates the centers of each cluster (assigning points to the closest center) until the solution becomes stable \cite{Jain}. Lloyd's algorithm is only guaranteed to reach a locally optimum solution and can be very sensitive to outliers and the initial points chosen \cite{Kanungo}.

The $k$-median objective function removes some worst-case behavior of $k$-means by restricting the cluster centers to be data points.  There are two common algorithms for this problem. The first is a local search algorithm called partition around medoids (PAM), which starts with $k$ randomly chosen data points as centers, and looks to swap one of the centers with another data point to improve the objective value. This repeats until no more improving swaps exist \cite{ESL}. The generalization of this algorithm which considers swaps of size $\leq p$ is a $(3 + \frac{2}{p})$-approximation algorithm \cite{Arya}.
Another common heuristic (called $k$-medoids throughout this paper), which is more computationally efficient than PAM \cite{Park}, is a variation on Lloyd's $k$-means algorithm.  Starting with randomly chosen centers, we assign each data point to its closest center, and then update the center within each cluster \cite{Park}. The algorithm repeats this until the solution becomes stable.

The $k$-median problem is also a fundamental problem in approximation algorithms where it is often presented in the context of opening service centers or factories to serve clients. However, these algorithms are often not implemented for clustering despite their improved guarantees on the solution. The best known approximation algorithm for this problem has an approximation guarantee of 2.592 and is produced using linear programming \cite{Wu}. A simpler LP rounding algorithm by Charikar and Li has a slightly worse guarantee of 3.25 but is easier to implement in practice \cite{Charikar}.

\subsection{Hierarchical Clustering}
The most common method for hierarchical clustering is agglomerative clustering. Agglomerative clustering starts with each data point in its own cluster, and proceeds to recursively merge the two closest clusters to form the next clustering level in the hierarchy, repeating until the entire data set is a single cluster.  There are several choices on how to measure distance in order to choose which clusters to merge. Single linkage (also called nearest neighbor) defines the distance between two clusters to be the minimum pairwise distance between the data points in each cluster.  Complete linkage clustering instead defines the distance to be the largest pairwise distance. A method between these two extremes is group average linkage which looks at the average pairwise distance between clusters. This strikes a balance between the closeness and compactness of the two techniques above and gives more reasonable results in practice \cite{ESL}. Lastly, there is Ward's method which chooses the two clusters to merge so as to minimize the sum of each cluster's variance \cite{Ward}.

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to greedily construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. Using the local $p$-swap $(3+\frac{2}{p})$-approximation algorithm mentioned above as the black-box $k$-median solver performed well in experiments on hierarchical $k$-median test problems \cite{Nagarajan}. While the authors tested the variants of this algorithm for different $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

@@

Recall the hierarchical $k$-median problem: given a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$, we represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$.
 A solution to the hierarchical $k$-median problem is a chain of nested solutions $(S_1, a_1) \subseteq (S_2, a_2) \subseteq \ldots \subseteq (S_l, a_l)$. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy.  
Consider any $k$-median algorithm $\mathcal{A}$. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce an augmentation step that nests these clusterings into the hierarchical framework and only increase costs by a bounded factor.

First, set $\beta = 3+\sqrt{3}$, and $\beta_0 = 1$ (for the randomized algorithm we switch $\beta = 6.355$ and set $\beta_0 = \beta^X$, where $X$ is chosen from a uniform distribution on $[0,1]$). Let $V_i$ be the solutions generated from running $\mathcal{A}$ for $k=n-i+1$ for $i=1, \ldots, n$. The cost of these solutions is determined by assigning each point to the nearest cluster center and is scaled so that the minimum cost for $k<n$ is at least one. We then bucket these solutions by cost into buckets of the form $[0,0], (\beta_0, \beta_0 \beta], (\beta_0 \beta, \beta_0 \beta^2], \ldots $.

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions: Set $S_1 = \overline{V}_1$ and $a_1$ to be the assignment that assigns each point to the closest point in $S_1$. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we construct $S_i$ by  adding the closest point $j \in S_{i-1}$ to $l$ for each $l \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

\begin{figure*}[!t]
\centering
\subfloat[Nested solution $(S_{i-1}, a_{i-1})$ ]{\includegraphics[width =1.75in]{Si-1}
\label{fig_Si-1}} \hspace{15mm}
\subfloat[Next set $\bar{V}_i$]{\includegraphics[width=1.75in]{Vi}
\label{fig_Vi}} \\
\subfloat[Next nested centers $S_i$]{\includegraphics[width=1.75in]{Si}
\label{fig_Si}}  \hspace{15mm}
\subfloat[Next nested solution $(S_i, a_i)$]{\includegraphics[width=1.75in]{Siai}
\label{fig_Siai}}
\caption{An example of nesting $\bar{V}_i$ with $(S_{i-1}, a_{i-1})$. Dark  nodes represent chosen centers and edges represent assignments to that center's cluster. First, node 2 chooses node from 1 from $S_{i-1}$ as its closest center and node 5 chooses node 4. Then, each cluster in $(S_{i-1}, a_{i-1})$ is assigned to the cheapest (lowest sum of distances) center.}
\label{fig_sim}
\end{figure*}

We then need to define an assignment $a_i$. For each $j \in X$ such that $a_{i-1}(j) \in S_{i}$, we set $a_i(j) = a_{i-1}(j)$. Otherwise, we set $a_i(j)$ to be the facility $i \in S_i$ such that assigning all points in $a_{i-1}^{-1}(j)$ (i.e. all points in $j$'s cluster in $S_{i-1}$) would have minimum cost. Using this assignment, $(a_i, S_i)$ is nested in $(a_{i-1}, S_{i-1})$. An example of this nesting is given in Fig.~\ref{fig_Si-1}-\ref{fig_Siai}.

@@

\subsection{Partitional Clustering Algorithms}
This nesting framework allows us to choose the partitional clustering algorithm $\mathcal{A}$. As mentioned above, PAM and $k$-medoids are two potential algorithms. PAM has the added benefit of being a 5-approximation algorithm for the partitional $k$-median problem which yields to an overall approximation ratio of $20.71 \cdot 5 = 103.55$ for the hierarchical $k$-median problem ($10.03 \cdot 5 = 50.15$ randomized). We looked at increasing the size of the local swaps in PAM to improve the approximation ratio but found that the approximation ratio of constructed solutions for given values of $k$ did not significantly improve while the computation time increased greatly.

We also consider three linear programming based algorithms, one of which has an approximation guarantee. We first present the linear programming formulation for the $k$-median problem. Suppose we have a set $X$. Let $s_i \in \{0,1\}$ be the variable representing whether or not we want to make $i$ a center for all $i \in X$. We also create a variable $a_{i,j} \in \{0, 1\}$ which represents whether or not we assign $i \in X$ to the cluster with center $j \in X$. Given this set and assignment, the $k$-median objective function is 
\begin{equation}
\sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j), 
\end{equation}
which we want to minimize. However, we are restricted by the fact that we can choose at most $k$ centers, every point must be assigned to a single cluster center, and that we can only assign a point to a center that is actually in $S$. This is represented by the following three equations, respectively:
\begin{eqnarray}
\sum_{i \in X} s_i   \leq k & ,\\
\sum_{j \in X} a_{i,j}  = 1 & \forall i \in X, \\
a_{i,j}   \leq s_j & \forall i, j \in X.
\end{eqnarray}
Therefore, solving the following integer program would directly find the optimal set of centers and assignment for the $k$-median problem.
\begin{eqnarray}
\text{Min } & \sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j) \\
\text{s.t. } & \sum_{i \in X} s_i   \leq k  \\
& \sum_{j \in X} a_{i,j}  = 1 \quad \forall i \in X \\
& a_{i,j}   \leq s_j \quad \forall i, j \in X \\
& s_i \in \{0,1\} \quad \forall i \in X \\
& a_{i,j} \in \{0,1\} \quad \forall i, j \in X
\end{eqnarray}
Relaxing the constraints so that $s_i \in [0,1]$ for all $i \in X$ and $a_{i,j} \in [0,1]$ for all $i,j \in X$ yields the linear programming (LP) relaxation for the $k$-median problem, which we can solve in polynomial time.

Interestingly, the solution to the linear programming is often integer valued \cite{Nagarajan}. Based on this fact, we propose two LP-based clustering algorithms: one that solves the LP relaxation and constructs $S$ by taking the points in $X$ with the $k$ highest $s_i$ values and one that solves the LP relaxation and constructs $S$ by choosing $i$ to be in $S$ with probability $s_i$, sampling for each $i \in X$ until $k$ centers are chosen. We refer to these algorithms as LP-greedy and LP-rand, respectively.
Charikar and Li \cite{Charikar} introduce a 3.25-approximation algorithm that uses this LP solution to smartly construct a set of centers. We refer to this algorithm as LP-Charikar. We choose to implement this LP algorithm rather than the LP-based approximation algorithm with best approximation ratio because the algorithm is simpler to implement (and we expect it to have a lower running time).

The algorithm works by first choosing a subset of data points $C \subseteq X$ to represent the demand. For each data point $i$, let $d_{av}(i) = \sum_{j \in X : a_{i,j} > 0} s_j \cdot d(i,j)$. We start with $C = \emptyset$. At each step, we choose a point $j \in X$ to add to $C$, and all points $j'$ such that $d(j,j') \leq 4 d_{av}(j') $ are removed from $X$. This is continued until $X$ is empty.

@@

All tests were conducted using Python 2.7.9 on a 64-bit Windows platform with a 4 core 3.4 Ghz Intel i7 processor and 16 GB physical memory. All code and the full computational results are available at \url{https://github.com/apaul29/Hierarchical-Code}.
We chose to test our methods on data from a few different contexts:
\begin{LaTeXdescription}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing $k$ random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] \cite{Beasley} From a collection of classic test data sets for approximation algorithms created by Beasley with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. This data is unlabeled.
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{LaTeXdescription}

@@

\subsection{Partitional Algorithms}
 For the black-box $k$-median algorithm, we implemented $k$-medoids, PAM, LP-greedy, LP-rand, and LP-Charikar in Python using Gurobi 6.0.3 to solve the linear programming instances. 
To avoid cycling, we limited the number of iterations in $k$-medoids to 1000. For PAM, we also needed to limit the number of swaps to achieve a reasonable runtime. Instead of searching over all potential swaps, we sample at most $n$ potential ones. We also ensure that some minimum progress is achieved by ensuring the algorithm decreases in objective by at least a factor of 0.9 in each iteration. At the end of PAM, we run 10 iterations of $k$-medoids to ensure the algorithm finds a local optimum. These choices were made by looking at the relative performance and computation time on random Gaussian data not in the data sets above.

@@

\subsection{Hierarchical Results}
After testing the partitional algorithms above, we implemented the nesting hierarchical framework, offering options for both the randomized and deterministic versions. For PAM and $k$-medoids, when solving for multiple values of $k$, we use the solution with $k$ centers as a warm start for $k+l$ centers by choosing an additional random $l$ centers. For the LP algorithms, Gurobi uses a warm start from solving the LP for $k+l$ to solve it for $k$ because we found this reverse direction to be faster in practice.

We also tested a more computationally friendly heuristic of the hierarchical framework which only buckets solutions for values of $k$ that are a power of two. This yields many fewer $k$-median instances to solve, reducing the run-time significantly. This was motivated by the fact that the difference between various $k$-median solutions will be larger when $k$ is small so we expect our sampling to still represent much of the underlying structure. For PAM, we only tested the logarithmic heuristic since, as also indicated by the partitional results, the run time for the full version proved impractical.

@@

There were several interesting trends in the performance. First, the deterministic versions had higher FScore measures and lower approximation coefficients on average compared to their randomized counterparts so we would advise to just use the deterministic version despite its higher approximation guarantee. Similarly, our simple LP heuristics performed comparably to the more advanced LP-Charikar algorithm. Also worth noting is the fact that the approximation ratio for $k$-medoids improved as $n$ increased in the $p$-median instances, which is surprising the values of $k$ for some of the instances increased proporitional to $n$ so we would expect the overall difficulty to be higher.

In terms of the computation time, the agglomerative clustering methods were significantly faster than all of nesting algorithms, which we would expect given how many $k$-median instances they have to solve. $k$-medoids was the fastest method, but $k$-medoids, LP-greedy, and LP-rand all performed around 4 orders of magnitude slower than the agglomerative methods (this order increased to 5 for a few of the very large $p$-median instances). LP-Charikar was the slowest method and was often at least 5 orders of magnitude slower than the agglomerative methods.

>>>>

\begin{abstract}
[TODO]
Hierarchical clustering organizes large data into smaller, more informative groups and offers a global view of the data structure by producing a hierarchy of clusters starting from individual data points and merging towards the cluster of all data. Typically, hierarchical clustering is done by greedily either merging or dividing clusters, called agglomerative and divisive clustering, respectively. Lin et al.~\cite{Lin} proposed a nesting framework that produces a hierarchal clustering from partitional clusterings for every possible number of clusters. This algorithm was introduced as an approximation algorithm for the hierarchical $k$-median problem and guarantees the resulting hierarchical layers are a limited distance from the original partitional clusters. However, this algorithm has not been previously placed in a clustering context. We test this algorithm using various methods for producing the partitional clusterings against the traditional hierarchical clustering methods and show both the approximation ratio and FScore measure improve significantly. We also propose a heuristic for this nesting framework which improves the run time and only suffers a slight decrease in approximation and FScore.
\end{abstract}

\section{Introduction and Notation}
% As the data we have access to grows in variety and size and becomes cheaper to store, we need methods to efficiently learn from this data. 
% Partitioning data (clustering) into similar groups (clusters) is one of the key tools to learn about a set of unlabeled data such as a set of customers who have visited a website or a set of handwritten documents.
% Clustering in these situations helps us learn underlying structure and classifications of our data, and it allows large data sets to be presented in a simplified and compressed view \cite{Jain}.

[INTRODUCE HIERARCHICAL CLUSTERING]

@@

Clustering algorithms date back more than 50 years, and there have been many variations introduced since then \cite{Jain}. However, clustering algorithms are often formulated as NP-hard problems with non-convex objective functions, which are inherently hard to solve \cite{Jain}.
One common objective function for partitional clustering is to minimize the sum of distances from each point to the median point of its cluster, called $k$-median or $k$-medoids clustering. More specifically, suppose we are given a set $X$ and a distance (dissimilarity) metric $d(i,j)$ for all $i,j \in X$. The goal is to find a set of centers $S \subseteq X$ such that $|S| \leq k$ so as to minimize 
\[ f(S) = \sum_{ i\in X} \min_{j \in S} d(i,j). \]
The clustering is obtained from the set of centers $S$ by assigning each data point to its nearest center.  An $\alpha$-approximation algorithm for this problem produces a set $S$ in polynomial time such that $f(S) \leq \alpha \cdot f(O)$ for all $O \subseteq X$ with $|O| = k$.

\begin{figure*}[!]
\centerline{\subfloat[Partitional Clustering]{\includegraphics[width =1.75in]{partitional_clustering}
\label{fig_part}}
\hfil
\subfloat[Hierarchical Clustering]{\includegraphics[width=2.5in]{hier_clustering}
\label{fig_hier}}}
\caption{A partitional clustering for $k=3$, where the colors represent the different partitions, and a dendrogram representation of a hierarchical clustering of the data, where edges represent mergeing of clusters.  Note the clustering on the third level (represented by the red line) corresponds to the partitional clustering on the left.}
\end{figure*}

@@

\subsection{Partitional Clustering}
Arguably the most common partitional clustering method is $k$-means, which chooses $k$ cluster centers in space so as to minimize the sum of squared distances from each point to its nearest cluster center \cite{Jain}.  $k$-means is usually solved using Lloyd's algorithm (also referred to simply as the $k$-means algorithm), which iteratively greedily updates the centers of each cluster (assigning points to the nearest center) until the solution becomes stable \cite{Jain}. Lloyd's algorithm is only guaranteed to reach a locally optimum solution and can be very sensitive to outliers and the initial points chosen \cite{Kanungo}.

The $k$-median objective function removes some worst-case behavior of $k$-means by restricting the cluster centers to be data points.  Two common algorithms for this problem are partition around medoids (PAM), and $k$-medoids. PAM is a local seach algorithm which starts with $k$ randomly chosen data points as centers, and looks to swap one of the centers with another data point to improve the objective value. This repeats until no more improving swaps exist \cite{ESL}. The generalization of this algorithm which considers swaps of size $\leq p$ is a $(3 + \frac{2}{p})$-approximation algorithm \cite{Arya}.
$k$-medoids, which is more computationally efficient than PAM \cite{Park}, is a variation on Lloyd's $k$-means algorithm.  Starting with randomly chosen centers, we assign each data point to its nearest center, and then update the center within each cluster \cite{Park}. The algorithm repeats this until the solution becomes stable.

The $k$-median problem is also a fundamental problem in approximation algorithms, where it is often presented in the context of opening service centers or factories to serve clients. However, these algorithms are often not implemented for clustering despite their improved guarantees on the solution. The best known approximation algorithm for this problem has an approximation guarantee of 2.592 and is produced using linear programming \cite{Wu}. A simpler LP rounding algorithm by Charikar and Li has a slightly worse guarantee of 3.25 but is easier to implement in practice \cite{Charikar}.

\subsection{Hierarchical Clustering}
The most common method for hierarchical clustering is agglomerative clustering. The agglomerative clustering procedure starts with each data point in its own cluster, and recursively merges the two nearest clusters to form the next clustering level in the hierarchy, repeating until the entire data set is a single cluster.  There are several choices on how to measure distance in order to choose the nearest clusters to merge. Single linkage (also called nearest neighbor) defines the distance between two clusters to be the minimum pairwise distance between the data points in each cluster.  Complete linkage clustering instead defines the distance to be the largest pairwise distance. A method between these two extremes is group average linkage which looks at the average pairwise distance between clusters. This strikes a balance between the closeness and compactness of the two techniques above and gives more reasonable results in practice \cite{ESL}. Lastly, there is Ward's method which chooses the two clusters to merge so as to minimize the sum of each cluster's variance \cite{Ward}.

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. Their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to greedily construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. Using the local $p$-swap $(3+\frac{2}{p})$-approximation algorithm mentioned above as the black-box $k$-median solver performed well in experiments on hierarchical $k$-median test problems \cite{Nagarajan}. While the authors tested the variants of this algorithm for different $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

@@

Recall the hierarchical $k$-median problem: given a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$, we represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$.
 A solution to the hierarchical $k$-median problem is a chain of nested solutions $(S_1, a_1) \subseteq (S_2, a_2) \subseteq \ldots \subseteq (S_n, a_n)$. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy.  
Consider any $k$-median algorithm $\mathcal{A}$. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce an augmentation step that nests these clusterings into the hierarchical framework and only increase costs by a bounded factor.

First, set $\beta = 3+\sqrt{3}$, and $\beta_0 = 1$ (for the randomized algorithm we instead let $\beta = 6.355$ and set $\beta_0 = \beta^X$, where $X$ is drawn uniformly from $[0,1]$). Let $V_i$ be the solutions generated from running $\mathcal{A}$ for $k=n-i+1$ for $i=1, \ldots, n$. The cost of these solutions is determined by assigning each point to the nearest cluster center and is scaled so that the minimum cost for $k<n$ is at least one. We then bucket these solutions by cost into buckets of the form $[0,0], (\beta_0, \beta_0 \beta], (\beta_0 \beta, \beta_0 \beta^2], \ldots $.

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions: Set $S_1 = \overline{V}_1$ and $a_1$ to be the assignment that assigns each point to the nearest point in $S_1$. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we construct $S_i$ by  adding the nearest point $j \in S_{i-1}$ to $l$ for each $l \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

\begin{figure*}[!t]
\centering
\subfloat[Nested solution $(S_{i-1}, a_{i-1})$ ]{\includegraphics[width =1.75in]{Si-1}
\label{fig_Si-1}} \hspace{15mm}
\subfloat[Next set $\bar{V}_i$]{\includegraphics[width=1.75in]{Vi}
\label{fig_Vi}} \\
\subfloat[Next nested centers $S_i$]{\includegraphics[width=1.75in]{Si}
\label{fig_Si}}  \hspace{15mm}
\subfloat[Next nested solution $(S_i, a_i)$]{\includegraphics[width=1.75in]{Siai}
\label{fig_Siai}}
\caption{An example of nesting $\bar{V}_i$ with $(S_{i-1}, a_{i-1})$. Dark  nodes represent chosen centers and edges represent assignments to that center's cluster. }
\label{fig_sim}
\end{figure*}
% First, node 2 chooses node from 1 from $S_{i-1}$ as its nearest center and node 5 chooses node 4. Then, each cluster in $(S_{i-1}, a_{i-1})$ is assigned to the cheapest (lowest sum of distances) center.

We then need to define an assignment $a_i$. For each $j \in X$ such that $a_{i-1}(j) \in S_{i}$, we set $a_i(j) = a_{i-1}(j)$. Otherwise, we set $a_i(j)$ to be the point $i \in S_i$ such that assigning all points in $a_{i-1}^{-1}(j)$ (i.e. all points in $j$'s cluster in $S_{i-1}$) would have minimum cost. Using this assignment, $(a_i, S_i)$ is nested in $(a_{i-1}, S_{i-1})$. An example of this nesting is given in Fig.~\ref{fig_Si-1}-\ref{fig_Siai}.

@@

\subsection{Partitional Clustering Algorithms}
Since the described nesting framework allows us to choose the partitional clustering algorithm $\mathcal{A}$, the choice of this algorithm is critical. PAM has the added benefit of being a 5-approximation algorithm for the partitional $k$-median problem, which yields to an overall approximation ratio of $20.71 \cdot 5 = 103.55$ for the hierarchical $k$-median problem ($10.03 \cdot 5 = 50.15$ randomized). We looked at increasing the size of the local swaps in PAM to improve the approximation ratio but found that the approximation ratio of constructed solutions for given values of $k$ did not significantly improve while the computation time increased greatly.

We also consider three linear programming based algorithms, one of which has an approximation guarantee. We first present the integer programming formulation for the $k$-median problem. Suppose we have a set $X$. Let for all $i \in X$, $s_i \in \{0,1\}$ be variables representing whether or not we want to make $i$ a center. We also create variables $a_{i,j} \in \{0, 1\}$ which represent whether or not we assign $i \in X$ to the cluster with center $j \in X$. Given this set and assignment, and distance metric $d$, the $k$-median objective function is 
\begin{equation}
\sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j), 
\end{equation}
which we want to minimize. However, we are restricted to choosing at most $k$ centers, every point must be assigned to a single cluster center, and we can only assign a point to another point that has actually been chosen as a center. These restrictions are represented by the following constraints, respectively:
\begin{eqnarray}
\sum_{i \in X} s_i   \leq k & ,\\
\sum_{j \in X} a_{i,j}  = 1 & \forall i \in X, \\
a_{i,j}   \leq s_j & \forall i, j \in X.
\end{eqnarray}
Therefore, solving the following integer program would directly find the optimal set of centers and assignment for the $k$-median problem.
\begin{eqnarray}
\text{Min } & \sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j) \\
\text{s.t. } & \sum_{i \in X} s_i   \leq k  \\
& \sum_{j \in X} a_{i,j}  = 1 \quad \forall i \in X \\
& a_{i,j}   \leq s_j \quad \forall i, j \in X \\
& s_i \in \{0,1\} \quad \forall i \in X \\
& a_{i,j} \in \{0,1\} \quad \forall i, j \in X.
\end{eqnarray}
Relaxing the constraints to $s_i \in [0,1]$ for all $i \in X$ and $a_{i,j} \in [0,1]$ for all $i,j \in X$ yields the linear programming (LP) relaxation for the $k$-median problem, which can be solved in polynomial time.

Interestingly, the solution to the LP relaxation is often integer valued \cite{Nagarajan}. In light of this, we propose two LP-based clustering algorithms that both first solve the LP relaxation.  The first proposed algorithm constructs the set of centers $S$ by taking the points in $X$ with the $k$ highest $s_i$ values.  The second constructs $S$ by choosing $i$ to be in $S$ with probability $s_i$, sampling for each $i \in X$ until $k$ centers are chosen. We refer to these algorithms as LP-greedy and LP-rand, respectively.

Charikar and Li \cite{Charikar} introduce a 3.25-approximation algorithm that uses LP relaxation solution to smartly construct a set of centers. We refer to this algorithm as LP-Charikar. We chose to implement this LP algorithm, rather than the LP-based approximation algorithm with best approximation ratio, because LP-Charikar is simpler to implement, and expected to have a lower running time.

[MAYBE CUT THE ALGORITHM DESCRIPTION AND SAY TO REFER TO THE ORIGINAL PAPER?]
The algorithm first chooses a subset of data points $C \subseteq X$ to represent the demand. For each data point $i$, let $d_{av}(i) = \sum_{j \in X : a_{i,j} > 0} s_j \cdot d(i,j)$. We start with $C = \emptyset$. At each step, we choose a point $j \in X$ to add to $C$, and all points $j'$ such that $d(j,j') \leq 4 d_{av}(j') $ are removed from $X$. This is continued until $X$ is empty.

@@

All tests were conducted using Python 2.7.9 on a 64-bit Windows platform with a 4 core 3.4 Ghz Intel i7 processor and 16 GB physical memory. All code and the full computational results are available at \url{https://github.com/apaul29/Hierarchical-Code}.
We tested the algorithms on the following data sets:
\begin{LaTeXdescription}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing $k$ random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] \cite{Beasley} From a collection of classic test data sets for approximation algorithms created by Beasley with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. This data is unlabeled.
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{LaTeXdescription}

@@

\subsection{Partitional Algorithms}
For $k$-median algorithms, we implemented $k$-medoids, PAM, LP-greedy, LP-rand, and LP-Charikar in Python, using Gurobi 6.0.3 to solve the linear programs. 
To avoid cycling, we limited the number of iterations in $k$-medoids to 1000. For PAM, we also limit the number of swaps to achieve a reasonable runtime. Instead of searching over all potential swaps, we sample at most $n$ potential swaps. We also ensure that some minimum progress is achieved by ensuring the algorithm decreases in objective by at least a factor of 0.9 in each iteration. At the end of PAM, we run 10 iterations of $k$-medoids to ensure the algorithm finds a local optimum. These choices were made by looking at the relative performance and computation time on random Gaussian data not included in the data sets described above.

@@

\subsection{Hierarchical Results}
We implemented both the determinisic and randomized versions of the nesting hierarchical framework. For PAM and $k$-medoids, when solving for multiple values of $k$, we use the solution with $k$ centers as a warm start for $k+l$ centers by choosing an additional random $l$ centers. For the LP algorithms, Gurobi uses a warm start from solving the LP for $k+l$ to solve it for $k$ because we found this reverse direction to be faster in practice.

We also tested a more computationally friendly heuristic of the hierarchical framework which only buckets solutions for values of $k$ that are a power of two. This yields many fewer $k$-median instances to solve, reducing the run-time significantly. This was motivated by the fact that the difference between various $k$-median solutions will be larger when $k$ is small, so we expect our sampling to still represent much of the underlying structure. For PAM, we only tested the logarithmic heuristic since the run time for the full version proved impractical.

@@

There were several interesting trends in the performance. First, the deterministic versions had higher FScore measures and lower approximation coefficients on average compared to their randomized counterparts, suggesting usage of deterministic version in practice despite its higher approximation guarantee. Similarly, our relatively simpler LP heuristics performed comparably to the more advanced LP-Charikar algorithm. Also worth noting is the fact that the approximation ratio for $k$-medoids improved as $n$ increased in the $p$-median instances, which is surprising, as the values of $k$ for some of the instances increased proporitional to $n$, so we would expect the overall difficulty to be higher.

In terms of computation time, the agglomerative clustering methods were significantly faster than all of nesting algorithms, which is to be expected given the numerous $k$-median instances the nesting algorithms must solve. $k$-medoids, LP-greedy, and LP-rand all performed around 4 orders of magnitude slower than the agglomerative methods (this order increased to 5 for a few of the very large $p$-median instances), with $k$-medoids being the fastest. LP-Charikar was the slowest method and was often at least 5 orders of magnitude slower than the agglomerative methods.
